---
title: "Lecture 2: Conditional Probability and Independence"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Simulating Conditional Probabilities

Recall, one interpretation of _probability_ is long-term relative frequency. Suppose you repeat a random experiment a large number of time, say $N$, and observe $n$ number of event $A$ The relative frequency 
$$\frac{n}{N}$$ 
is approximately the probability $P(A)$.

### Example: Rolling an 8-Sided Fair Die Twice


$$A=\text{sum of the throws is divisible by 4}$$
$$B=\text{the two throws are the same}$$

Recall we can generate $N$ simulations of a single roll with:

```{r echo=TRUE, eval=FALSE}
roll <- sample(x = 1:8, size = N, replace = TRUE)

```

We can use 2 separate `sample` calls to mimic 2 rolls and then combine them using `cbind`.

```{r tworolls, exercise=TRUE, exercise.lines=7}
N <- 10 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)
print(tworolls)

```

Using the `N=1,000,000`, compute $P(A)$, $P(B)$, $P(A\cap B)$, $P(A|B)$, and $P(B|A)$.

$$P(A)$$

```{r pa, exercise=TRUE, exercise.lines=15}
N <- 10^6 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)

```

$$P(B)$$

```{r pb, exercise=TRUE, exercise.lines=15}
N <- 10^6 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)

```

$$P(A\cap B)$$

```{r paandb, exercise=TRUE, exercise.lines=15}
N <- 10^6 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)

```

$$P(A|B)$$

```{r pagivenb, exercise=TRUE, exercise.lines=15}
N <- 10^6 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)
# select those that satisfy b
tworolls_givenb <- tworolls[tworolls[,1] == tworolls[,2],]
  
```

$$P(B|A)$$

```{r pbgivena, exercise=TRUE, exercise.lines=15}
N <- 10^6 # number of simulations
roll1 <- sample(1:8, N, replace = TRUE)
roll2 <- sample(1:8, N, replace = TRUE)
# use `cbind` to combine them into a matrix 
tworolls <- cbind(roll1, roll2)

```


## Simulating Conditional Events

In this exercise, we will compute the probability of an event based on conditional probabilities.

### Example

Suppose the probability of an adult Canadian experiencing a regular flu during a winter is 12% without a vaccination and with a vaccination, the infection probability is reduced to 3%. If 35% of the adult population in Canada gets vaccinated every year, what is the probability that a randomly selected Canadian adult experiences a regular flu during a winter?


```{r flu, exercise=TRUE, exercise.lines=15}
N <- 100
# the probabilities
pvac <- 0.35
pflu_vac <- 0.03
pflu_no_vac <- 0.12
# simulate vaccinated population with TRUE / FALSE otherwise
s_vac <- sample(
  x = c(T, F), size = N, replace = TRUE,
  prob = c(pvac, 1 - pvac)) # we can pass the probabilities manually
# simulate flu infection based on vaccination
```

> Hint: After simulating vaccinated popoulations, you can use a for loop with an if / else statement to simulate flu infections. If you are familiar with `tidyverse` or `dplyr`, you can check out `rowwise` operations.
