---
title: "Lecture 2: Conditional Probability and Independence"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## R Preparation

### Matrix

We discussed vectors in last class. It is often useful to work in 2-dimensions
when working with data. In R, a 2-dimensional object consisting of values of a 
single data type.

You can create a matrix using `matrix(<vector of values>, <number of rows>)`.

```{r matrix-num}
matrix(1:9, 3)
```

```{r matrix-letters}
matrix(LETTERS, 2)
```

> `LETTERS` is a prepopulated vector variable with the capital alphabets.

You can access the individual elements using `[` like vectors. If you use a single
value or single vector, it treats the matrix as a vector concatenating the columns
in order.

```{r matrix-access-single}
mat_ex <- matrix(1:12, 3)
mat_ex
mat_ex[5] # 5th value
mat_ex[c(2, 11)] # 2nd and 11th values
```

More intuitively, you can use 2-dimensional coordinates to extract the elements
separated by a comma. The first value corresponds to the row and the second value
to the column. The index starts at 1 like a single value index.

```{r matrix-2d-coord}
mat_ex[2, 3] # second row, third column
mat_ex[c(1, 3), 3] # first and third row, third column
mat_ex[c(2, 3), c(1, 4)] # second and third row, first and fourth column
```

If you leave one dimension blank, you can extract across the whole rows or columns.

```{r matrix-blank-dimension}
mat_ex[1, ] # all values from the first row
mat_ex[ , 4] # all values from the fourth column
```

There are a few convenient functions provided for working with multidimensional
objects. For this exercise, you will need `rowSums()`. Check the documentation
for the function as well as related functions.

```{r rowSums-doc, exercise=TRUE}
```

### Relational Operators

+  To check whether two (vectors of) values are equation, you can use `==`.

```{r equal}
"a" == "a"
"a" == "A"
1:6 == (0:5 + 1)
1 == 1:5
```

> To check whether they are NOT equal you can use `!=`. 

+  To check whether a value or elements of a vector is found in another vector,
use `%in%`. 

```{r in}
"a" %in% letters
1:5 %in% 2:3
1:5 %in% 5:15
```

> `letters` is a prepopulated vector variable with the lowercase alphabets.

+  You can also test for values being greater/less than (or equal to) similarly

```{r gteqlteq}
5 <= 5
5 < 5:10
1:5 > 3:7
"a" > "b" # alphabetical order
TRUE > FALSE # TRUE is 1, FALSE is 0
```

### Logical Operators

+  `&&` is AND for operations between single logicals.

```{r and-single}
TRUE && TRUE
TRUE && FALSE
1 && 1 # you can also turn non-zeros/0 to T/F
```

+  `&` is AND for both single values and vectors.

```{r}
TRUE & FALSE
c(TRUE, TRUE, TRUE) & c(TRUE, FALSE, TRUE)
```

+  `||` is OR for operations between single logicals.

```{r or-single}
TRUE || TRUE
TRUE || FALSE
0 || 2 # you can also turn non-zeros/0 to T/F
```

+  `|` is OR for both single values and vectors.

```{r}
TRUE || FALSE
c(TRUE, TRUE, FALSE) | c(TRUE, FALSE, FALSE)
```

+ ! negates the immediately following logical value / expression.

```{r negate}
TRUE && ! FALSE
! TRUE || FALSE
! 1 || 2 & 0
```


You may also find `all()` and `any()` useful. See their documentations.

```{r all-any-doc, exercise=TRUE}
```

### Practice

In the following code chunk's `mat_practice`,

1. find the row numbers where the values in the row sum up to smaller than 15
2. find the column numbers where the values in the column sum up to larger than 10
3. check whether all values in the second row are larger than values in the fourth row AND 
value 5 is in the last column


```{r practice-logical-relational, exercise=TRUE, exercise.lines=10}
mat_practice <- matrix(sample(1:24, 24), 6)
mat_practice
1:6[ ] # fill in the bracket with a logical statement
1:4[ ] # fill in the bracket with a logical statement
```

## Estimating Conditional Probabilities

We will practice estimating conditional probabilities in multiple experiment setting.

### Example: Rolling Two Fair Dice

You roll two fair dice.

$A$ is the event that _sum of the rolls is divisible by 4_,  
$B$ is the event that _the two roll are the same_.

Let's estimate $P(A|B)$ with $N$ simulations.

1. Simulating $N$ simulations of multiple experiments.

Recall, to simulate a single roll:

```{r single-roll, eval=FALSE}
sample(1:6, 1)
```

You may choose to write a loop and sample single roll twice in each iteration.

```{r loop-single-rolls, eval=FALSE}
roll1 <- numeric(N)
roll2 <- numeric(N)
for (i in 1:N) {
  roll1[i] <- sample(1:6, 1)
  roll2[i] <- sample(1:6, 1)
}
```

Or, you can avoid writing a loop by sampling $4\times N$ _independent_ rolls and
placing them into a matrix. 

```{r N, include=FALSE}
N <- 10
```

```{r all-rolls}
m <- 2 # number of rolls in each simulation
rolls <- sample(1:6, N * m, replace = TRUE) # replace = TRUE makes each roll independent
mat <- matrix(rolls, nrow = N) # place them into matrix
mat
```

> The columns represent the rolls and the rows represent the simulations.

2. Extract the values you need to identify the events.

We can use `rowSums()` to sum the values of each row.

```{r sum-apply}
rowSums(mat)
```

To access all simulations of the first rolls, you can use

```{r first-roll}
mat[, 1]

```

and for the second rolls

```{r second-roll}
mat[ , 2]
```


3. Identify the simulations that satisfy the conditional event and compute the estimated probability.

Complete the code chunk below to estimate $P(A|B)$ with $N$ simulations.

```{r compute-conditional-p, exercise=TRUE, exercise.lines=15}
N <- 1000
m <- 2
rolls <- sample(1:6, N * m, replace = TRUE) 
mat <- matrix(rolls, nrow = N)
sum_of_rolls <- rowSums(mat)
first_rolls <- mat[ , 1]
second_rolls <- mat[ , 2]
# complete the logic here
pr_a_given_b
```


## Simulating Conditional Events

In this exercise, we will simulate evens based on conditional probabilities.

### Example: Flu in Canada

Suppose the probability of an adult Canadian experiencing a regular flu during 
a winter season is 12% without a vaccination and with a vaccination, the 
infection probability is reduced to 5%. If 35% of the adult population in Canada 
gets vaccinated every year, what is the probability that a randomly selected 
Canadian adult experiences a regular flu during a winter?


```{r flu, exercise=TRUE, exercise.lines=30}
N <- 1000
# the probabilities
p_vac <- 0.35
p_flu_vac <- 0.03
p_flu_no_vac <- 0.12
# simulate vaccinated population with TRUE / FALSE otherwise
s_vac <- sample(
  x = c(T, F), size = N, replace = TRUE,
  prob = c(p_vac, 1 - p_vac) # we can adjust the probabilities 
)
flu <- logical(N)
# simulate flu infection based on vaccination in a loop
p_flu
```

> Hint: After simulating vaccinated popoulations, you can use a for loop with an if / else statement to simulate flu infections. 
