---
title: "Lecture 2: Conditional Probability and Independence"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Simulating Conditional Events

In this exercise, we will compute the probability of an event based on conditional probabilities.

Suppose the probability of an adult Canadian experiencing a regular flu during a winter season is 12% without a vaccination and with a vaccination, the infection probability is reduced to 3%. If 35% of the adult population in Canada gets vaccinated every year, what is the probability that a randomly selected Canadian adult experiences a regular flu during a winter?


```{r flu, exercise=TRUE, exercise.lines=30}
N <- 100
# the probabilities
pvac <- 0.35
pflu_vac <- 0.03
pflu_no_vac <- 0.12
# simulate vaccinated population with TRUE / FALSE otherwise
s_vac <- sample(
  x = c(T, F), size = N, replace = TRUE,
  prob = c(pvac, 1 - pvac)) # we can pass the probabilities manually
# simulate flu infection based on vaccination
```

> Hint: After simulating vaccinated popoulations, you can use a for loop with an if / else statement to simulate flu infections. If you are familiar with `tidyverse` or `dplyr`, you can check out `rowwise` operations.

